---
title: "Pedotransfer functions for estimating carbon and organic matter content in Rio Grande do Sul, Brazil"
author: "Alessandro Samuel-Rosa, Douglas Adams Weiler, Sandro Giacomini"
date: "12 October 2017"
bibliography: "~/Dropbox/jabref/biblio.bib"
csl: pesquisa-agropecuaria-brasileira.csl
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load packages
library(dplyr)
library(magrittr)
library(latticeExtra)
```

# Abstract {-}

**Index terms**:

# Introduction

There exist several analytical methods to measure the content of carbon in soil samples. The standard method 
for measuring the *total carbon* content is the dry combustion method, which is based on the oxidation of 
both organic and inorganic forms of carbon at very high temperatures (> 1000°C) in the presence of oxygen and 
copper oxide [@NelsonEtAl1996; @PansuEtAl2006]. However, wet digestion methods, which are based on the 
reduction of Cr~2~O~7~^2-^ by the organic compounds present in soil samples, are the most commonly used 
[@PansuEtAl2006; @DonagemaEtAl2011]. This is mainly due to three reasons. First, the implementation and 
maintenance costs associated with the use of the standard method are often high, what makes it unsuitable for 
small laboratories. Second, provided that the quantity of inorganic forms of carbon in soil samples is 
negligible, the accuracy of wet digestion methods is comparable to the standard method [@RheinheimerEtAl2008, 
@GattoEtAl2009]. Last, most soil classification and fertilizer recommendation systems were built using criteria 
based on soil *organic carbon* and/or *organic matter* content as estimated using wet digestion methods 
[@IUSSWorkingGroupWRB2015; @SantosEtAl2013a; @SilvaEtAl2016].

Wet digestion methods have long been criticized because the ion Cr^6+^ is toxic to all forms of life 
[@RodellaEtAl1994; @PimentelEtAl2006]. As such, efforts have been made to stimulate the use of alternative 
methods. For example, the loss on ignition method, which is based on the quantification of the amount of mass 
lost by soil samples after oxidation of organic compounds at relatively high temperatures (> 300ºC) in the 
presence of oxygen [@PansuEtAl2006]. In Brazil, this method was pointed many times as being an attractive 
alternative to measure the soil organic matter content -- as used in fertilizer recommendation systems -- due 
to the very low environmental hazards associated with it use [@BrunettoEtAl2006; @EscosteguyEtAl2007]. With 
respect to the measurement of the organic carbon content, despite the standard dry combustion method being a 
natural alternative [@GattoEtAl2009], its widespread usage in Brazil has been hindered by the reasons mentioned
above. Thus, alternative wet digestion methods that employ less Cr~2~O~7~^2-^ have been proposed to try to
minimize their well known environmental hazards [@RheinheimerEtAl2008]. These efforts helped shaping a complex 
scenario in which disparate analytical methods are employed in different studies and soil laboratories.

Comparing and using soil carbon data produced in different studies that employed disparate analytical methods
require the development of harmonization mechanisms. Harmonization consists of converting the values of a soil
property produced using an arbitrary analytical method to values that "look like" the values that would be 
produced if the reference analytical method had been used instead [@BatjesEtAl2017]. For example, the Van 
Bemmelen factor, 1.724, which relies on the strong assumption that soil organic matter is composed of 58% of 
carbon, is commonly used to transform organic carbon content to organic matter content [@Pribyl2010]. Similar 
"conversion factors" have already been estimated in Brazil, for example, to enable the adoption of the loss 
on ignition method as a substitute for the wet digestion method used to measure the organic matter content 
[@EscosteguyEtAl2007], and to convert the organic carbon values produced by wet digestion methods to values of 
the standard dry combustion method [@PereiraEtAl2006; @RheinheimerEtAl2008; @GattoEtAl2009].

In a broader sense, a conversion factor can be understood as being a pedotransfer function. A pedotransfer
function consists of an empirical model that can be used to predict the values of a soil property from other 
soil properties that are easier or cheaper to measure or that are more readily available [@McBratneyEtAl2002]. 
Applied to the context of soil carbon data, understanding conversion factors as being pedotransfer functions 
means taking the carbon content measured with disparate analytical methods as if they are different soil 
properties -- which is specially true for total carbon, organic carbon, and organic matter content. These 
empirical models generally take the form of statistical models such as a linear regression. Conversion factors 
themselves are linear regression models based on the strong assumption that the regression line passes 
through the origin [@Pribyl2010].



and that standard significance tests are resistant to the violation of the iid-normal 
assumption [@Webster2001; ].

have already developed such linear regression models to predict soil carbon or organic matter content under 
various soil conditions, specially with varying granulometry, taxonomy, and land use [@BrunettoEtAl2006; @PereiraEtAl2006; @GattoEtAl2009; @SampaioEtAl2012].



, thus enabling estimating the most likely value of a soil property along with an uncertainty measure, e.g. the prediction interval.


A few studies were developed in Brazil in the last decade in order to compare different methods and find proper pedotransfer functions -- more specificaly correction factors -- to the standard method. However, none of them compared the wet digestion, the loss on ignition, the routine method used in laboratories under ROLAS (wet digestion) and the standard dry combustion method at once. Besides, a few issues related to the representativeness of the datasets and to statistical assumptions appear in all studies.

The loss on ignition method was evaluated as a substitute for the wet digestion and routine method by @BrunettoEtAl2006. However, they used only *n* = 19 soil samples from the Serra Gaúcha region of the State of Rio Grande do Sul (RS). The clay content of the samples ranges between 210 and 530 g kg^-1^ and the organic matter content ranges between 2.9% and 13.3%. Besides, the data has an asymmetric distribution, with most data points concentrated bellow the value of 8% of organic matter. In similar study @EscosteguyEtAl2007 evaluated the loss on ignition method as a substitute for the routine method, using *n* = 19 soil samples from eleven soil classes of the RS (clay content ranging between 89 and 729 g kg^-1^) and *n* = 49 soil samples (sandy soils) from the Central Depression region of the RS. The organic matter content of the samples is bellow 5.5% and the data has a normal distribution.

The precision of the wet digestion method in estimating the organic matter content of soils measured by the dry combustion method was evaluated by @RheinheimerEtAl2008. They used *n* = 18 soil samples from a Ferralsol from the Planalto region of the RS and *n* = 75 soil samples from an Acrisol from the Central Depression region of the RS. The first set has a clay content ranging between 352 a 425 g kg^-1^ and the second set has a clay content ranging between 95 a 468 g kg^-1^. The organic matter content of the samples ranges between 0.3% and 7.0% and the data has an asymmetric distribution. In order to establish pedotransfer functions to estimate the organic matter content of soils from different climatic conditions under Eucaliptus plantation, @GattoEtAl2009 compared the wet digestion with the dry combustion method. They used *n* = 99 soil samples from three soil classes collected in the Centro-Leste region of the State of Minas Gerais. No information is given about the clay content and mineralogy of the soils. The organic matter content in soils is bellow 4.0% and the data has a normal distribution.

The present study takes place due to the poor representativeness and failure in attending to statistical assumptions of datasets used in previous studies and the need for more information about the loss on ignition method. The primary objective is to evaluate the use of three methods (wet digestion, loss on ignition and the routine method) to estimate the organic carbon/matter content of soils measured by the dry combustion method. Second, to evaluate the loss on ignition method as a substitute method for the routine method used in laboratories under ROLAS.

# Material and Methods

## Soil samples description and preparation

```{r, echo=FALSE, message=FALSE}
camada <- readr::read_csv('../data/camada.csv', locale = readr::locale(decimal_mark = ','), comment = 'unidade')
observacao <- readr::read_csv('../data/observacao.csv', locale = readr::locale(decimal_mark = ','), na = '-')
observacao <-
  observacao %>% 
  mutate(uso_id = gsub('Floresta', 'Native forest', .data$uso_id)) %>% 
  mutate(uso_id = gsub('Vegetação secundária', 'Secondary vegetation', .data$uso_id)) %>% 
  mutate(uso_id = gsub('Agricultura', 'Crop agriculture', .data$uso_id)) %>% 
  mutate(uso_id = gsub('Silvicultura', 'Forestry', .data$uso_id)) %>% 
  mutate(uso_id = gsub('Campo nativo', 'Animal husbandry', .data$uso_id))
```

We evaluated the methods of estimating the organic carbon/matter content in *n* = 105 air-dried < 2 mm-soil samples collected at 27 sites in the State of Rio Grande do Sul (RS), Brazil. The samples cover eleven soil groups (Acrisols, Alisols, Cambisols, Ferralsols, Fluvisols, Gleysols, Leptosols, Nitisols, Planosols, Regosols, and Vertisols) derived from several parent materials (sedimentary, volcanic, and metamorphic). The clay content of the soil samples varies between 40 and 800 g kg^-1^ (Figure \@ref(fig:clay-content)). The mineralogical assemblage is diverse (quartz, iron oxide, and 2:1 and 1:1 phyllosilicates). Several land use types are represented: rangeland, farmland, fallow land, plantation (Eucalyptus sp.) and native forests. The climate of the sampling sites varies from humid subtropical to temperate climate, both without a defined dry season. Mean annual precipitation across RS ranges from approximately 1300 to 2000 mm. The mean annual temperature across RS is about 18°C, whichever is colder in higher sites (> 800 m a.s.l.) and warmer in lower sites (< 200 m a.s.l.). Soil samples collection procedures are described elsewhere [@CasaliEtAl2006, Barros (2007), @DickEtAl2008, @Britzke2010, @PotesEtAl2010, Vieira (2010), @Samuel-RosaEtAl2013a, @Samuel-RosaEtAl2013]. We grounded the soil samples in an agate mortar (< 1 mm) and stored them in microtubes (Eppendorf type) in the dark at room temperature until analysis.

```{r clay-content, echo=FALSE, fig.cap='Key characteristics of the soil samples, such as the clay content and sampling depth, and their sampling sites, such as the soil classification and type of land use and occupation.', fig.width=14, fig.height=14}
p1 <- histogram(
  ~ argila_naoh_pipeta, camada, xlab = 'Clay content (g/kg)', ylab = 'Percent of total', col = 'lightgray')
p2 <- histogram(
  ~ prof_sup, camada, xlab = 'Sampling depth (cm)', ylab = 'Percent of total', col = 'lightgray')
p3 <- barchart(
  observacao$taxon_sibcs_2009, xlab = 'Soil classification', ylab = 'Percent of total', horizontal = FALSE, 
  col = 'lightgray')
p4 <- barchart(
  observacao$uso_id, xlab = 'Land use and ocupation', ylab = 'Percent of total', horizontal = FALSE, col = 'lightgray')
gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## Analytical methods

### Total carbon by dry combustion

Sample aliquots of 50 to 60 mg [ESTES VALORES ESTÃO CORRETOS???] in titanium capsules were used to estimate the total soil organic carbon content in an elementary CHNS analyzer (FlashEA 1112, Thermo Finnigan, Milan, Italy).

### Organic carbon by wet digestion

Sample aliquots of 0.050 to 0.500 g were placed in glass digestion tubes (80 mL). The amount of sample used varied according to the SOM content estimated through visual interpretation of soil color. The correlation between SOM content and sample aliquot used as calculated in the input training dataset is r = -0.8436. Every digestion tube received an aliquot of 10 mL of 0.067 mol L-1 sulfocromic solution [K2Cr2O7 + H2SO4] and a small reflux funnel to avoid loss of reagent during digestion. A digestion block with capacity for 40 samples was used: 36 tubes with soil sample + 3 tubes with blank + 1 tube with sulfuric acid [H~2~SO~4~] and a thermometer for temperature measurement. Digestion at 150ºC last 30 min. Three blanks were prepared and set aside at room temperature to estimate the loss of reagent due to heat in the digestion block. After digestion the tubes were set aside at room temperature to cool down. Next, we transfered the solution to Erlenmeyer flasks (250 mL) with 60 mL of distilled water and 2 mL of concentrated orthophosphoric acid [H~3~PO~4~] and 3 drops of 1% diphenylamine. The solution was titrated using 0.1 mol L^-1^ ammonium ferrous sulfate [Fe(NH~4~)~2~(SO~4~])~2~.6H~2~O] until persistent green color. This method was originaly adapted from @Mebius1960 and @YeomansEtAl1988.

### Organic matter by wet digestion

Adapted from @TedescoEtAl1995. Sample aliquots of 2.5 mL were placed in Erlenmeyer flasks (50mL). Every Erlenmeyer flask received an aliquot of 15 mL of 0.067 mol L^-1^ sulfocromic solution [Na~2~Cr~2~O~7~ + H~2~SO~4~]. The flasks were heated in a water bath at 75 to 80°C during 30 min and shaken for 5 min. A water aliquot of 15 mL was added to the flask and let overnight (15 to 18 hours). In the next day, we sampled an aliquot of 3.0 mL to a small cup with 3.0 mL of distilled water. The supernatant absorbance was measured at 645 nm. We weighted 2.5 mL of each soil sample to calculate the sample density and transform the results to a weight basis.

### Organic matter by loss on ignition

Sample aliquots of 300 to 500 mg in hand-made aluminum capsules (20 mg) (oven dried at 360°C) were placed in an oven at 105°C for 24 hours. We varied the sample aliquot used due to sample availability what had no noticeable effect on the analytical results as showed by linear correlation analysis (data not shown). Next, we placed the oven-dried samples in a desiccator containing silica gel and weighted after they had cooled down. The amount of mass lost between the first and second weightings was used to calculate the water content in the sample. Next, we placed the samples in an oven at 360ºC for 2 hours. Next, samples were placed in a desiccator containing silica gel and weighted after they had cooled down. The organic matter content in the samples was calculated from the mass lost between the second and third weightings. This method was adapted from @SchulteEtAl1996.

## Data analysis

```{r carbon-content, echo=FALSE, fig.cap='Carbon and organic matter content in soil samples according to the four analythical methods.', fig.width=14, fig.height=14}
p1 <- histogram(
  ~ carbono_dry, camada, xlab = 'Total carbon content by dry combustion (g/kg)', ylab = 'Percent of total', col = 'lightgray')
p2 <- histogram(
  ~ carbono_wet, camada, xlab = 'Organic carbon content by wet digestion (g/kg)', ylab = 'Percent of total', col = 'lightgray')
p3 <- histogram(
  ~ mo_wet, camada, xlab = 'Organic matter content by wet digestion (g/kg)', ylab = 'Percent of total', col = 'lightgray')
p4 <- histogram(
  ~ mo_loi, camada, xlab = 'Organic matter content by loss on ignition (g/kg)', ylab = 'Percent of total', col = 'lightgray')
gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

# Results and Discussion

# References